Taxonomic bias can create spurious results in microbiome differential-abundance analyses unless it is properly accounted for

### 2021-07-12 article.Rmd
Import comments from Google Doc

@@

@@

# Bias can cause spurious results in proportion-based DA analysis

@@

@@

Figure \@ref(fig:vaginal-example) illustrates the effects of taxonomic bias on the measured proportions for hypothetical vaginal samples from a single woman at two points in time, during which the community shifts from being dominated by *Lactobacillus* to being dominated by *Gardnerella*.
In each case, we observe a greater proportion of *Lactobacillus* than is truly present, and less of the other two species (Figure \@ref(fig:vaginal-example)A).
This positive multiplicative error arises because *Lactobacillus* is more efficiently measured than the average species in each sample.
@mclaren2019cons showed that the observed and actual proportions of a taxon $i$ in a sample $s$ are related by the equation
\begin{align}
  (\#eq:error)
  \text{observed}_{i}(s) = \text{actual}_{i}(s) \cdot \frac{\text{efficiency}_{i}}{\text{mean efficiency}(s)}
\end{align}
where $\text{mean efficiency}(s)=\text{actual}_j(s)*\text{efficiency}_j$ in the denominator is the average efficiency of all taxa in that sample.
Intuitively, Equation \@ref(eq:error) says that the taxon will be overrepresented by the extent to which its measurement efficiency is greater than the average efficiency of cells in the sample.
Figure \@ref(fig:vaginal-example)B shows the efficiencies of the three taxa and the mean efficiency in each sample.
The (multiplicative) difference between the taxon efficiencies and the mean efficiency determines the (multiplicative) error in each sample; for example, the mean efficiency in the second sample is 2.5, leading to a 29/2.5=11.6X error in the proportion of *Lactobacillus*.
The error in the second sample is larger—specifically, by 6X—than the first sample because the high proportion of *Gardnerella* has reduced the sample mean efficiency (and thus increased the efficiency of *Lactobacillus* relative to the mean) by 6X.

@@

@@

@@

@@

# Implications for real-world inference

@@

@@

@@

@@

@@

@@

So far we have ignored error associated with our absolute abundance measurements.
In fact, these measurements can also have a tendency to measure contributions from some taxa more efficiently than others.
**TODO: State how this could create problems.
Might also note the ambiguity in what the truth/target is (e.g.
live cells, or all cells; cells, or 16S rRNA gene copies; observations in Jian, Salonen, and Korpela 2021).**
Interestingly, taxonomic bias may actually make qPCR a better method for absolute quantification than cell counting for 16S rRNA gene sequencing experiments.
16S rRNA gene qPCR measures the concentration of 16S rRNA gene copies in the extracted DNA and is therefore necessarily affected by three large sources of bias in 16S rRNA gene experiments: extraction, amplification, and copy-number variation.
Yet these biases are shared by the sequencing measurement.
Though they make qPCR measurements a bad proxy for total cell density, when used for absolute differential abundance inference the shared bias in the qPCR and sequencing measurements can cancel, leaving relative accurate fold changes.
(Jian, Salonen, and Korpela 2021 suggests some of these ideas and APPENDIX gives a mathematical justification.)
If common primers are used by qPCR and 16S rRNA gene sequencing, the bias due to primer mismatches and other sources of variation in amplification could even be accounted for in this manner.
Yet ideally we would still be able to estimate and correct remaining error due to unshared bias or to mechanisms, such as saturation during DNA extraction, that could break the assumed proportionality between qPCR measurements and total abundance.

@@

@@

@@

Because targeted absolute-abundance measurements are expected to provide (relatively) accurate  estimates of fold changes, they can be used to validate and even correct the DA results derived from a proportion-based absolute DA analysis.
We illustrate the basic idea using the @lloyd2020evid experiment described above.
Absolute abundances in this study were estimated by multiplying MGS proportions by total abundance measured by cell counting; for comparison, qPCR was used to measure specific microbial taxa.
Because the error in fold changes or a regression from the MGS measurements is taxon-independent, the difference between the two methods for these reference taxa informs us about the difference for all taxa.
@lloyd2020evid found close agreement for the specific taxa also measured by qPCR (Table 1), suggesting that variation in sample mean efficiency did not significantly distort their results.
In principle, a joint statistical analysis of all measurements would allow inferring variation in the sample mean efficiency across samples and obtaining calibrated fold change and regression estimates for all taxa [APPENDIX].

>>>>

- On whether MGS can be quantitative
    - Karen Lloyd: Is there a reference that explains the disagreement? In my field, it hasn't really been explicitly addressed adequately.
    - Michael McLaren: I'll have to look/think about this some more. I was trying to express a vague impression I have based on various things I've read and heard, and how community sequencing is used in practice, that implicitly suggest there is disagreement or at least a tension in the field where on the one hand there are many papers on the problems with community sequencing and on the other hand many papers (often by the same authors, and including me and Ben) using it to make quantitative inferences
    - Karen Lloyd: You could reference Mallick et al., 2017 DOI 10.1186/s13059-017-1359-z; Or Barlow et al. 2020 https://doi.org/10.1038/s41467-020-16224-6

@@

- 'many hope' - consider make small tweaks to phrases like this to sound less critical

@@

## Notes for revision

- introduce differential abundance as the focus earlier
- improve introduction of absolute abundance vs. relative abundance, and methods for obtaining absolute abundances
    + might include computational methods in scope, and use the term "calibration"; and ref to the future appendix that surveys the various methods
    + consider giving general explanation and/or specific examples of why absolute abundances are important
# Bias can cause spurious results in proportion-based DA analysis

@@

- TODO: Add main-message sentence; see Ben's note below

@@

- Ben flagged that we want to make sure that it is clear from the text that this example is not just a one-off case of bias mattering, but is an illustration of a general/universal problem. The later discussion helps with this already; but what will help more is to start this section with a main-message sentence stating the general problem first.

Figure \@ref(fig:vaginal-example) illustrates the effects of taxonomic bias on the measured proportions for hypothetical vaginal samples from a single woman at two points in time, during which the community shifts from being dominated by *Lactobacillus* to being dominated by *Gardnerella*.
In each case, we observe a greater proportion of *Lactobacillus* than is truly present, and less of the other two species (Figure \@ref(fig:vaginal-example)A).
This positive multiplicative error^[what is the right terminology for multiplicative error that is greater than 1?] arises because *Lactobacillus* is more efficiently measured than the average species in each sample.
@mclaren2019cons showed that the observed and actual proportions of a taxon $i$ in a sample $s$ are related by the equation
\begin{align}
  (\#eq:error)
  \text{observed}_{i}(s) = \text{actual}_{i}(s) \cdot \frac{\text{efficiency}_{i}}{\text{mean efficiency}(s)}
\end{align}
where $\text{mean efficiency}(s)=\text{actual}_j(s)*\text{efficiency}_j$ in the denominator is the average efficiency of all taxa in that sample.
Intuitively, Equation \@ref(eq:error) says that the taxon will be overrepresented by the extent to which its measurement efficiency is greater than the average efficiency of cells in the sample.
Figure \@ref(fig:vaginal-example)B shows the efficiencies of the three taxa and the mean efficiency in each sample.
The (multiplicative) difference between the taxon efficiencies and the mean efficiency determines the (multiplicative) error in each sample; for example, the mean efficiency in the second sample is 2.5, leading to a 29/2.5=11.6X error in the proportion of *Lactobacillus*.
The error in the second sample is larger—specifically, by 6X—than the first sample because the high proportion of *Gardnerella* has reduced the sample mean efficiency (and thus increased the efficiency of *Lactobacillus* relative to the mean) by 6X.

- see note on terminology for mult. error

@@

- on the description of competition - Ben suggested adding "for reads in the sequencer", but this addition doesn't quite make sense the way I've written this sentence, which is talking about the mathematical operation of normalization to proportions. But this paragraph deserves more attention, as the fact that we are imposing competition  via normalization to proportions, but the experiment also imposes competition that we can't avoid by just not normalizing, is subtle

@@

- TODO: Add strike-throughs to the above equation
- Revision should hone the selection of equations

@@

- on "(e.g., sediment depth)": Want to confirm with KL that I'm understanding this correctly. She notes: "Even though these values are sequential, we use them as discrete variables because the samples are only from one depth each. I'm not sure if this distinction matters here." To which I replied: "right....you first converted sediment depth to time, and then regressed log abundance on time to estimate growth rate? (new line) Here I was just using sediment depth as an example variable, but I could replace with time to more match your analysis"
- Consider different term than 'slope' since it's not a slope in the binary case

@@

- Note about this figure: It is currently framed as O = A - \bar B to show the effect of error on inferred DA results; it could be flipped to A = O + \bar B to show how a correction might be done.
# Implications for real-world inference

@@

- Discussion of the vaginal example in the above paragraph should be distilled in conjunction with the discussion of this example in the previous section

@@

- In last sentence, insert the the values of the LFCs (ideally with evaluated R code)

@@

- on the example of 'ocean sediment' in the above P: I asked KL "is it reasonable to characterize the ocean sediment communities from your 2020 paper as "high diversity"? In particular, species- or OTU-level Inverse Simpson values are what's relevant here. I'm guessing it's intermediate diversity - much less than gut and soil, but much more than the vaginal microbiome", and she replied "The total microbial diversity in marine sediments worldwide is similar to that of soil. I'll email you Hoshino et al., 2020 which addresses this." However I'm not sure if the high diversity KL is noting applies to individual soil samples (i.e. what scale)
- Should make sure the above paragraph's claims are precise and accurate; see AW's feedback
- "As diversity increases" specifically refers to inverse simpson diversity / q=2. Should I explicitly state that?

@@

- mclaren2019cons measures overall extraction efficiency, not lysis efficiency; is that important for the above text?

@@

- first sentence in above P: "the abundances and efficiencies" - the idea is that both are determined by genetically encoded traits
- this is prob a tricky idea for readers and will have to take care in communicating it, and find nice refs

@@

- Ben flags that this concept/section main idea should be highlighted more

So far we have ignored error associated with our absolute abundance measurements.
In fact, these measurements can also have a tendency to measure contributions from some taxa more efficiently than others.
**TODO: State how this could create problems.
Might also note the ambiguity in what the truth/target is (e.g.
live cells, or all cells; cells, or 16S rRNA gene copies; observations in Jian, Salonen, and Korpela 2021).**
Interestingly, taxonomic bias may actually make qPCR a better method for absolute quantification than cell counting for 16S rRNA gene sequencing experiments.
16S rRNA gene qPCR measures the concentration of 16S rRNA gene copies in the extracted DNA and is therefore necessarily affected by three large sources of bias in 16S rRNA gene experiments: extraction, amplification, and copy-number variation.
Yet these biases are shared by the sequencing measurement.
Though they make qPCR measurements a bad proxy for total cell density^[See note from KL below], when used for absolute differential abundance inference the shared bias in the qPCR and sequencing measurements can cancel, leaving relative accurate fold changes.
(Jian, Salonen, and Korpela 2021 suggests some of these ideas and APPENDIX gives a mathematical justification.)
If common primers are used by qPCR and 16S rRNA gene sequencing, the bias due to primer mismatches and other sources of variation in amplification could even be accounted for in this manner.^[See discussion below]
Yet ideally we would still be able to estimate and correct remaining error due to unshared bias or to mechanisms, such as saturation during DNA extraction, that could break the assumed proportionality between qPCR measurements and total abundance.

- from KL on qPCR being a bad proxy: We showed this empirically for all published measurements in marine sediments in Lloyd et al., 2013 doi:10.1128/AEM.02090-13. Apologies for the self-citation.
- In the above sentence "If common primers are used...", KL asked a question that shows I'm not being clear enough about what I'm talking about here. KL: "Wouldn't you want to use taxa specific primers for qPCR and general primers for the amplicon libraries to do this analysis? If they're the same primer sets, then it doesn't tell you anything about the relative abundance of individual taxa within the population grabbed by the primers." my reply: "The "solutions" section talks about using targeted qPCR; this section is talking about qPCR with "universal" primers for the purpose of estimating the total absolute abundance, for use in FRAxC instead of cell counting. There have been quite a few papers doing this recently and 16S sequencing companies are starting to offer it as a service so I've been thinking through the implications. This section needs to be expanded and the point I'm making about using qPCR for FRAxC is still somewhat preliminary and something we should chat about"

@@

- from gdoc: this distinction of ratios vs proportions needs to be spelled out more explicitly; could even use my seminar figure (or equations) to help
- later thoughts: This distinction is so central for how I understand and explain the key ideas, should refer to it early and often

@@

- note from KL on choice of reference taxon: "It seems like you'd have to be pretty careful to pick a reference taxon that was not changing a lot between samples."
    - this comment indicates I need to be clearer about the three reference strategies - assume constant AA; (but may not be available so can) spike-in at constant AA; (but that might not be convenient, so can) measure AA to adjust for changes in AA.
    - my reply in the doc: "the idea is that either the reference is a taxon spiked at a fixed abundance to each sample, or it is a taxon you've measured independently with ddPCR or qPCR and so can account for its variation through the given equation." "There are also computational methods that attempt to identify taxa that don't vary (under the assumption that most taxa don't vary), which is an approach inspired from the RNAseq/gene expression literature (where it is probably a more valid assumption that most genes have consistent expression); Ben in his comments was suggesting we mention this approach too. "
- On taxa not competing in targeted measures: "I think DNA-based measurements would still be affected by competition if DNA extraction yield is a strongly saturating function of input. But if this saturation isn't very strong then the effect should remain much less than the MGS measurement, which also has competition imposed by various normalization steps "
- on 'considerations' noted in the last sentence: "perhaps most notable is how to deal with a given reference taxon not being present or detectable in every sample "

@@

Because targeted absolute-abundance measurements are expected to provide (relatively) accurate  estimates of fold changes, they can be used to validate and even correct the DA results derived from a proportion-based absolute DA analysis.
We illustrate the basic idea using the @lloyd2020evid experiment described above.
Absolute abundances in this study were estimated by multiplying MGS proportions by total abundance measured by cell counting; for comparison, qPCR was used to measure specific microbial taxa.^[see note]
Because the error in fold changes or a regression from the MGS measurements is taxon-independent, the difference between the two methods for these reference taxa informs us about the difference for all taxa.
@lloyd2020evid found close agreement for the specific taxa also measured by qPCR (Table 1), suggesting that variation in sample mean efficiency did not significantly distort their results.
In principle, a joint statistical analysis of all measurements would allow inferring variation in the sample mean efficiency across samples and obtaining calibrated fold change and regression estimates for all taxa [APPENDIX].

- 'specific microbial taxa' - I'm not sure if/how to address the taxonomic levels issue
- Details: Table 1 indicates that the doubling time estimated for Bathyarchaeota in Core 30 was 10.1 by FRAxC and 10.3 by qPCR, which corresponds to slopes (log abundance / time) of 1/10.1 and 1/10.3. The difference between these slopes, 1/10.3 - 1/10.1, is an estimate of the slope of the mean efficiency and hence is the correction that should be applied to all taxa’s FRAxC estimates.
